---
title: "Lab 5 - Working with NEON Metagenome Assembled Genomes"
author: "Katelynn Burns"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  freeze: auto
---

## On the Computer

```{r}
library(tidyverse)
library(knitr)
library(janitor)
library(DT)
```

## kable

```{r}
iris_setosa <- iris %>% 
filter(Species == "setosa") %>% 
filter(Sepal.Length > 5)
```

```{r}
library(knitr)
kable(iris_setosa)
```

## DT

```{r}
library(DT)
datatable(iris_setosa)
```

```{r}
datatable(
  iris %>% 
    filter(Species == "setosa") %>% 
    filter(Sepal.Length > 5)
)
```

```{r}
iris_setosa <- iris %>% 
  filter(Species == "setosa") %>% 
  filter(Sepal.Length > 5)

datatable(iris_setosa)
```

## Examples using the NEON data table

```{r}
NEON_MAGs <- read_tsv("/work/pi_bio678_umass_edu/data_NEON/exported_img_bins_Gs0166454_NEON.tsv")
```

```{r}
head(NEON_MAGs)
```

```{r}
str(NEON_MAGs)
```

```{r}
NEON_MAGs <- NEON_MAGs |> janitor::clean_names()
```

```{r}
NEON_MAGs %>% 
  count(bin_quality, sort = TRUE) 
```

```{r}
kable(
  NEON_MAGs %>% 
   count(bin_quality) 
)
```

```{r}
datatable(
  NEON_MAGs%>% 
    filter(bin_quality == "HQ")
)
```

```{r}
kable(
NEON_MAGs %>% 
  select(c(gtdb_taxonomy_lineage, total_number_of_bases)) %>% 
  filter(total_number_of_bases > 10000000)
)
```

```{r}
datatable(
NEON_MAGs %>% 
  filter(str_detect(gtdb_taxonomy_lineage, 'Bacteroidota'))
)
```

```{r}
datatable(
NEON_MAGs %>% 
  filter(str_detect(genome_name, 'Yellowstone NP'))
)
```

```{r}
NEON_MAGs_tax <- NEON_MAGs %>% 
  separate(gtdb_taxonomy_lineage, c("domain", "phylum", "class", "order", "family", "genus"), "; ", remove = FALSE) 
```

```{r}
datatable(
  NEON_MAGs_tax %>% 
    count(phylum, sort = TRUE)
)
```

```{r}
NEON_MAGs_tax_sample <- NEON_MAGs_tax %>% 
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("genome_name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(genome_name, c("site","sample_name"), " - ") %>% 
  # Get rid of the the common string "S-comp-1"
  mutate_at("sample_name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(sample_name, c("site_ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("subplot", "layer", "date"), "-") 
```

```{r}
datatable(
  NEON_MAGs_tax_sample %>% 
    count(site, sort = TRUE)
)
```

## Exercises

### Exercise 1

Use view(iris) to see the whole data table. Subset the table based on a different species than was used in the example. Display the table using DT::datatable.

```{r}
view(iris)
```

```{r}
versicolor_data <- subset(iris, Species == "versicolor")
```

```{r}
datatable(versicolor_data)
```

### Exercise 2

Display using DT::datatable the NEON MAGs from the individual assemblies that have at least 1 16S rRNA.

```{r}
NEON_MAGs |>
  filter(x16s_r_rna >= 1) |>
  datatable()
```

### Exercise 3

Display a table of the MAGs from Lower Tombigbee with only the columns for the genome_name, gtdb_taxonomy_lineage, and estimated MAG genome size (total_number_of_bases).

```{r}
lower_tombigbee_MAGs <- NEON_MAGs %>%
  filter(str_detect(genome_name, 'Lower Tombigbee'))%>%
  select(genome_name, gtdb_taxonomy_lineage, total_number_of_bases)
DT::datatable(lower_tombigbee_MAGs)
```

### Exercise 4

Display a table with the Class counts at Lyndon B. Johnson National Grasslands.

```{r}
LBJ_class_counts <- NEON_MAGs_tax %>%
  filter(str_detect(genome_name, 'Lyndon B. Johnson National Grassland')) %>%
  count(class, sort = TRUE)
DT::datatable(LBJ_class_counts)
```

### Exercise 5

Display a table with the counts for the Phylum Actinomycetota at each Site.

```{r}
actino_counts <- NEON_MAGs_tax %>%
  filter(phylum == "Actinomycetota") %>%
  count(genome_name, sort = TRUE)
DT::datatable(actino_counts)
```
